Does high personal relevance mitigate the persuasion of symbolic elements in government public communications?

### 2018-06-21 paper3_relevance and symbols_01.Rmd
Add files via upload

---
title: "Does high personal relevance mitigate the effect of symbolic elements in government public communications?"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Previous public administration studies have examined whether public communications can affect citizens trust in government organizations and their policies. Recent experimental studies have demonstrated that these communications can enhance trust not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations (e.g. logos, images and celebrities). Still, under which conditions are citizens more/less likely to be emotionally persuaded by these symbols? Building on the social psychology Elaboration Likelihood Model (ELM), I theorize that citizens are less susceptible to symbols when they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. I put this theory to test through a survey experiment, where I examine circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens' trust in the Environment Protection Ministry's policies for reducing the air pollution in a specific area (the Haifa-Bay) and reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The experimental results do not support the hypothesis that higher personal relevance attenuates the effect of symbols. They tentatively suggest that symbols are more likely to persuade citizens who have predominant negative attitudes.*

@@

While the abovementioned studies have substantiated the distinct effect of symbols, they did not theorize, nor empirically examine the conditions under which are these symbols are likely to be more/less effective in persuading citizens. The research gap has great significance. From the organizations' practical point of view, ...

Employing a broader, democratic normative stance, we should be also concerned from the effect of symbols. Symbolic elements can elicit undue trust in poorly performing organizations and in ill-conceived policies. Accordingly, it may enable public organizations to escape justified public criticism [@alonbarkat_2017]. Hence, by understanding when citizens are more likely to be affected by symbols, we  may enable us to identify the

Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *under what conditions do symbolic elements  increase citizens' trust in governemnt policy?* And specifically, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. I expect that this effect is attenuated the more citizens perceive the communications as personally relevant, because they are more motivated to think about their messages. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow() %>% round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions). In addition, I examine the moderating role of the perceived personal relevance of these policies.

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I did not find support for the hypothesis that the effect of the symbolic elements is attenuated by the perceived personal relevance of the policy issue. I discuss the theoretical and normative implications of these findings.

@@

>*H<sub>1</sub> - Symbolic elements in public organizations' communications are likely to increase citizens' trust in the organizations and in the policies presented in the communications.*

>*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

>*H<sub>3</sub> - Symbolic elements are likely to decrease the amount of thinking about the content of the communication.*

>*H<sub>4</sub> - The abovementioned conditioned effect of symbolic elements on citizens' trust is likely to be mediated by the decrease in the amount of thinking (a moderated mediation hypothesis).*

>>>>

---
title: "Does high personal relevance mitigate the persuasion of symbolic elements in government public communications?"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

*Previous public administration studies have examined whether public communications can affect citizens trust in government organizations and their policies. Recent experimental studies have demonstrated that these communications can enhance trust not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations (e.g. logos, images and celebrities). Still, under which conditions are citizens more/less likely to be emotionally persuaded by these symbols? Building on the social psychology Elaboration Likelihood Model (ELM), I theorize that citizens are less susceptible to symbols when they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. I put this theory to test through a survey experiment, where I examine circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens' trust in the Environment Protection Ministry's policy plans for reducing the air pollution in a specific area (the Haifa-Bay) and reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The experimental results do not support the hypothesis that higher personal relevance attenuates the effect of symbols. They tentatively suggest that symbols are more likely to persuade citizens who have predominant negative attitudes.*

@@

While the abovementioned studies have substantiated the distinct effect of symbols, they did not theorize, nor they empirically examined the conditions under which are these symbols are likely to be more/less effective in persuading citizens. This research gap has great significance. From the practical point of view of public organizations, this may enable them to design more effective communications which may enhance citizens' collaboration with their policies. Employing a broader perspective of normative democratic theory, the emotive effect of symbols can be perceived as problematic. It stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Morover, as argued by @alonbarkat_2017, symbolic communication may elicit undue trust in poorly performing organizations, and enable public organizations to escape justified public criticism. Hence, by understanding when citizens are more/less likely to be affected by symbols, we may be able to better assess the potential risk for governments' manipulative use of symbolic communications.

To address this literature gap, I draw upon the social psychology Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011]. I hypothesize that citizens are less susceptible to symbols the more they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. Similarly, higher perceived personal relevance are expected to enhance citizens' inclincation to scrutinize the substantive content of communication. Employing a survey experiment, I test this hypothesis in the context of the Israeli Environment Protection Ministry, and its policy plans for reducing the air pollution in a specific area (the Haifa-Bay) and for reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others.

@@

*H<sub>1</sub> - Symbolic elements in communications are less (more) likely to increase citizens' trust in governemnt organizations and policicies, the more (less) citizens perceive the policy issue in the communication as having personal relevance for themselves.*

*H<sub>2</sub> - The logical quality of the communication is more (less) likely to affect citizens' trust in governemnt organizations and policicies, the more (less) citizens perceive the policy issue in the communication as having personal relevance for themselves.*
